---
title: "Assignment #2 — Statistical analysis of thermal comfort research"
author: "Urwa Irfan"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## Setup
```{r setup-packages}
library(tidyverse)
library(ggplot2)
library(gt)
library(ggh4x)
library(patchwork)
library(ggcorrplot)
```

```{r setup-theme}
theme_set(theme_light(base_size = 10))
theme_update(
  strip.background = element_rect(fill = "#F5F5F5", color = "grey"),
  strip.text = element_text(color = "black"),
  strip.text.y = element_text(angle = 90)
)
theme_update(panel.grid = element_line(linetype = "dotted"))
cal_palette2 <- c("#3B7EA1", "#C4820E")
cal_palette3 = c("#3b7ea1", "#b0750c", "#003262")
cal_palette4 <- c("#003262", "#b0750c", "#3b7ea1", "#fdb515")
cal_palette5 <- c("#003262", "#b0750c", "#3b7ea1", "#fdb515", "#D9661F")
load_palette = c("#010090", "#8f0300")
```

```{r setup-data}
load("data/arch241a2.rda")
d <- arch241a2
dat = tibble(d) |>
   rename(T_op = Top) |> # for clarity
   mutate(Subject = factor(Subject, levels = as.character(sort(as.numeric(levels(Subject)))))) # relevel so it's sorted numerically, but kept as character for plotting
```

## About this dataset — read before you start

Thirty subjects each sat through seven conditions in a climate chamber in
Guangzhou, giving 30 × 7 = 210 rows.

| Variable | Meaning | Notes |
|---|---|---|
| `Top` | Operative temperature, °C | Equals air temperature here (MRT = Ta) |
| `RH` | Relative humidity, % | Two levels: 50 and 70 |
| `Sex` | F / M | 15 subjects each |
| `Subject` | Subject ID, 1–30 | Each ID appears 7 times|
| `TS` | Thermal sensation | ASHRAE scale, −4 cold … 0 neutral … +4 hot |
| `TC` | Thermal discomfort | 0 = very comfortable … 4 = very uncomfortable |
| `TA` | Thermal acceptability | Higher = more acceptable |
| `PMV` | Predicted mean vote | Calculated from the design conditions |
| `PPD` | Predicted % dissatisfied | Calculated from PMV |
| `Experiment` | Condition A–F, plus P | P is pre-conditioning, not a test condition |

## 1. Statistical summary of all variables

```{r tbl-q1}
#| tbl-cap: "Statistical summary"

summarize_variable = function(df, variable) {
   df |>
      # select(.data[[variable]]) |>
      summarise(
         Minimum = min(.data[[variable]]),
         `25th Quartile` = quantile(.data[[variable]], 0.25),
         Median = median(.data[[variable]]),
         `75th Quartile` = quantile(.data[[variable]], 0.75),
         Maximum = max(.data[[variable]]),
         Mean = mean(.data[[variable]]),
         `Standard Deviation` = sd(.data[[variable]])
      ) |>
      mutate(variable = variable)
}

summary_results = tibble()
variables = names(dat)[5:9]

for (var in variables) {
   summary_results = rbind(summary_results, summarize_variable(dat, var))
}
tbl_q1 = summary_results |>
   relocate(variable) |>
   mutate(variable = case_when(
      variable == "TS" ~ "Thermal Sensation (TS)",
      variable == "TC" ~ "Thermal Comfort (TC)",
      variable == "TA" ~ "Thermal Acceptability (TA)",
      variable == "PMV" ~ "Predicted Mean Vote (PMV)",
      variable == "PPD" ~ "Percentage People Dissatisfied (PPD)",
   )) |>
   gt(rowname_col = "variable") |>
   fmt_number()

tbl_q1
gtsave(tbl_q1, "output/q1_statistical-summary.html")
# Think: does a mean make sense for a factor? For a design variable like Top?
```

**Interpretation.** [WRITE 2–3 SENTENCES. What would you put in a paper?]


## 2. Histograms of TC and TS

```{r fig-q2}
#| fig-cap: "Histograms"
#| fig-width: 6
#| fig-height: 3

plt_q2 = dat |>
   select(TC, TS) |>
   pivot_longer(cols = c(TC, TS), names_to = "variable", values_to = "value") |>
   mutate(variable = case_when(
      variable == "TS" ~ "Thermal Sensation (TS)",
      variable == "TC" ~ "Thermal Comfort (TC)")
   ) |>
   ggplot(aes(value)) +
      geom_histogram(binwidth = 1, color = "black", fill = cal_palette2[1]) +
      labs(x = "Vote", y = "Number of votes") +
      scale_y_continuous(breaks = seq(0, 120, 20)) +
      facet_wrap(~ variable, scales = "free_x") +
      facetted_pos_scales(
         x = list(
            scale_x_continuous(limits = c(0, 4)),
            scale_x_continuous(limits = c(-4, 4), breaks = seq(-4, 4, 1))
         )
      )

plt_q2
ggsave("output/q2_histograms.png")
```

**Interpretation.** [WRITE 2–3 SENTENCES]
i have simplified to single vote count rather than continuous scale


## 3. Boxplots of TS vs Sex, Subject and Experiment

```{r fig-q3}
#| fig-cap: "Box plots"
#| fig-width: 8
#| fig-height: 6

q3_boxplot = function(df, var) {
   df |>
   ggplot() +
      geom_boxplot(aes(.data[[var]], TS)) +
      labs(y = "Thermal Sensation (TS)") +
      scale_y_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4))
}

dat_q3 = dat |>
   filter(Experiment != "P") 
   
plt_q3_sex = q3_boxplot(dat_q3, "Sex")
plt_q3_subject = q3_boxplot(dat_q3, "Subject")
plt_q3_experiment = q3_boxplot(dat_q3, "Experiment")
plt_q3 = (plt_q3_sex + plt_q3_experiment) / plt_q3_subject

plt_q3
ggsave("output/q3_boxplots.png")
# The Subject one is the least obvious and the most informative — look at it carefully.
```

**Interpretation.** [WRITE 2–3 SENTENCES]


## 4. Correlation matrix

Variables: PMV, PPD, RH, TA, TC, TS, Top.

```{r fig-q4}
#| fig-cap: "Correlation matrix"
#| fig-width: 6
#| fig-height: 6

plt_q4 = dat |>
   select(T_op, RH, TS, TC, TA, PMV, PPD) |>
   cor() |>
   ggcorrplot(type = "lower", outline.color = "black", lab = TRUE) +
      labs(x = "", y = "", fill = "Correlation\ncoefficient") +
      scale_fill_gradient2(low = "#4c59ea", high = "#ea544c") +
      theme_minimal() +
      theme(legend.position = "inside", legend.position.inside = c(0.2, 0.8), legend.background = element_rect(colour = "#e0e0e0"),
         axis.text.x = element_text(angle = 0))

plt_q4
ggsave("output/q4_correlation-matrix.png")
```

```{r fig-q4_high-corr}
#| fig-cap: "Correlation matrix"
#| fig-width: 6
#| fig-height: 4

# dat |>
#    ggplot() +
#       geom_point(aes(T_op, PMV))

```

PMV/PPD were directly calculated from the environmental conditions.(it just so happens that PMV is directly proportional to T_op - check?)

**Interpretation.** [WRITE 2–3 SENTENCES]

Two of these correlations are strange. One pair is r = 1.00, and another pair
is deterministically related yet shows a weak r. Find them, plot them, and
explain what happened.


## 5. Normality of TC and TS

```{r q5}
# [WRITE YOUR CODE] A test AND a plot. Do not rely on the p-value alone.
```

**Interpretation.** [WRITE 2–3 SENTENCES. Are they normal? Does it matter, and for what?]


## 6. Does Sex significantly influence TS?

```{r q6}
# [WRITE YOUR CODE]
# Then answer this, in your text: Sex is a property of a SUBJECT, and you have
# 30 subjects but 210 rows. What is your real sample size, and what does using
# 210 do to your p-value?
# Try at least two approaches and report whether they agree.
```

**Interpretation.** [WRITE 2–3 SENTENCES]


## 7. Linear regression: TS as a function of Top

```{r q7}
# [WRITE YOUR CODE] Fit the model, then TEST ITS VALIDITY.
# The four assumptions: linearity, constant variance, normal residuals,
# independent observations. Check all four and say which ones hold.
```

**Interpretation.** [WRITE 2–3 SENTENCES. Report the slope with units, the R²,
and the neutral temperature — the Top at which predicted TS = 0. Compare it to
what PMV predicts.]


## 8. Your own research question

**Question:** [WRITE YOUR QUESTION]

```{r q8}
# [WRITE YOUR CODE]
```

**Interpretation.** [WRITE 2–3 SENTENCES — the take-away message you would put
in a paper.]


## Session information

```{r session}
sessionInfo()
```
