---
title: "Assignment #2 — Statistical analysis of thermal comfort research"
author: "Urwa Irfan"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## Setup
```{r setup-packages}
library(tidyverse) # dplyr and related packages
library(ggplot2) # plotting
library(gt) # printing tables
library(ggh4x) # plotting customisation
library(patchwork) # combining plots
library(ggcorrplot) # correlation matrix
library(rstatix) # statistical analysis
library(Metrics) # statistical analysis
library(car) # statistical analysis
```

```{r setup-theme}
theme_set(theme_light(base_size = 10))
theme_update(
  strip.background = element_rect(fill = "#F5F5F5", color = "grey"),
  strip.text = element_text(color = "black"),
  strip.text.y = element_text(angle = 90)
)
theme_update(panel.grid = element_line(linetype = "dotted"))
cal_palette2 <- c("#3B7EA1", "#C4820E")
cal_palette3 = c("#3b7ea1", "#b0750c", "#003262")
cal_palette4 <- c("#003262", "#b0750c", "#3b7ea1", "#fdb515")
cal_palette5 <- c("#003262", "#b0750c", "#3b7ea1", "#fdb515", "#D9661F")
load_palette = c("#010090", "#8f0300")
```

```{r setup-data}
load("data/arch241a2.rda")
d <- arch241a2
dat = tibble(d) |>
   rename(T_op = Top) |> # for clarity
   mutate(Subject = factor(Subject, levels = as.character(sort(as.numeric(levels(Subject)))))) |> # relevel so it's sorted numerically, but kept as character for plotting
   filter(Experiment != "P") # because this was not a real test condition
```

## 1. Statistical summary of all variables

```{r tbl-q1}
#| tbl-cap: "Statistical summary"

summarize_variable = function(df, variable) {
   df |>
      # select(.data[[variable]]) |>
      summarise(
         Minimum = min(.data[[variable]]),
         `25th Quartile` = quantile(.data[[variable]], 0.25),
         Median = median(.data[[variable]]),
         `75th Quartile` = quantile(.data[[variable]], 0.75),
         Maximum = max(.data[[variable]]),
         Mean = mean(.data[[variable]]),
         `Standard Deviation` = sd(.data[[variable]])
      ) |>
      mutate(variable = variable)
}

summary_results = tibble()
variables = names(dat)[5:9]

for (var in variables) {
   summary_results = rbind(summary_results, summarize_variable(dat, var))
}
tbl_q1 = summary_results |>
   relocate(variable) |>
   mutate(variable = case_when(
      variable == "TS" ~ "Thermal Sensation (TS)",
      variable == "TC" ~ "Thermal Comfort (TC)",
      variable == "TA" ~ "Thermal Acceptability (TA)",
      variable == "PMV" ~ "Predicted Mean Vote (PMV)",
      variable == "PPD" ~ "Percentage People Dissatisfied (PPD)",
   )) |>
   gt(rowname_col = "variable") |>
   fmt_number()

tbl_q1
gtsave(tbl_q1, "output/q1_statistical-summary.html")
```

**Interpretation.**
For this question, I excluded the independent and categorical variables (T_op, RH, Sex, Subject, Experiment), because I don't think a statistical analysis of these alone provides any insights about the research. The PMV and PPD were directly calculated from the independent variables, so their distribution is also not particularly helpful.

For the measured data collected (thermal acceptability, comfort, and sensation), we see a broad distribution within their ranges, with the mimimum/maximum values close to the extremes on the scales. The mean and median values are similar to each other (indicating the data may be normally distributed). The mean/median for TS and TA fall near the 'neutral' condition of 0, but the mean/median for TC is around 2 (slightly uncomfortable). This is an interesting incongruence in the data: if only the mean for TS/TA was near 0, I would guess that most occupants were always uncomfortably cold or hot with an even spread averaging out to 0. However, since the median is also near 0, perhaps it's due to the nature of the survey design and how occupants interact with the questions.

## 2. Histograms of TC and TS

```{r fig-q2}
#| fig-cap: "Histograms"
#| fig-width: 6
#| fig-height: 3

plt_q2 = dat |>
   select(TC, TS) |>
   pivot_longer(cols = c(TC, TS), names_to = "variable", values_to = "value") |>
   mutate(variable = case_when(
      variable == "TS" ~ "Thermal Sensation (TS)",
      variable == "TC" ~ "Thermal Comfort (TC)")
   ) |>
   ggplot(aes(value)) +
      geom_histogram(binwidth = 1, color = "black", fill = cal_palette2[1]) +
      labs(x = "Vote", y = "Number of votes") +
      scale_y_continuous(breaks = seq(0, 120, 20)) +
      facet_wrap(~ variable, scales = "free_x") +
      facetted_pos_scales(
         x = list(
            scale_x_continuous(limits = c(1, 4), expand = expansion(mult=0.2)),
            scale_x_continuous(limits = c(-4, 4), breaks = seq(-4, 4, 1))
         )
      )

plt_q2
ggsave("output/q2_histograms.png", width = 6, height = 3, units = "in")
```

**Interpretation.**
While data for TC and TS were collected on a continuous scale, I have simplified the bins on the histograms to single vote counts for clearer interpretation. The histograms help clarify the previous summary: TS does look mostly normally distributed, but TC is skewed toward lower values. The TS data also largely agrees with the TC data: most people did not experience extreme thermal sensations (±3/4 bars are small) and thus were not extremely uncomfortable.

## 3. Boxplots of TS vs Sex, Subject and Experiment

```{r fig-q3}
#| fig-cap: "Box plots"
#| fig-width: 8
#| fig-height: 6

q3_boxplot = function(df, var) {
   df |>
   ggplot() +
      geom_boxplot(aes(.data[[var]], TS)) +
      labs(y = "Thermal Sensation (TS)") +
      scale_y_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4))
}

plt_q3_sex = q3_boxplot(dat, "Sex")
plt_q3_subject = q3_boxplot(dat, "Subject")
plt_q3_experiment = q3_boxplot(dat, "Experiment")
plt_q3 = (plt_q3_sex + plt_q3_experiment) / plt_q3_subject

plt_q3
ggsave("output/q3_boxplots.png", width = 8, height = 6, units = "in")
```

**Interpretation.** Thermal sensation does not seem to vary significantly by sex, but has a broad distribution across the experiments and subjects. It makes sense that the thermal sensation generally increases as the experimental operative temperatures were increased from A-F; we can also infer that relative humidity seems to have a much smaller impact than temperature, because the C & D and E & F box plots are very similar. Most subjects have median TS close to 0 with two notable exceptions (16 and 19) who also reported some of the largest ranges. There are only two outlier points (as `ggplot` defines them, values more than 1.5xIQR from the 1st & 3rd quartiles), with agrees with the previous charts in that most people don't experience extreme thermal sensations.

## 4. Correlation matrix

```{r fig-q4}
#| fig-cap: "Correlation matrix"
#| fig-width: 6
#| fig-height: 6

# Variables: PMV, PPD, RH, TA, TC, TS, Top.
plt_q4 = dat |>
   select(T_op, RH, TS, TC, TA, PMV, PPD) |>
   cor() |>
   ggcorrplot(type = "lower", outline.color = "black", lab = TRUE) +
      labs(x = "", y = "", fill = "Correlation\ncoefficient") +
      scale_fill_gradient2(low = "#4c59ea", high = "#ea544c") +
      theme_minimal() +
      theme(legend.position = "inside", legend.position.inside = c(0.2, 0.8), legend.background = element_rect(colour = "#e0e0e0"),
         axis.text.x = element_text(angle = 0))

plt_q4
ggsave("output/q4_correlation-matrix.png", width = 6, height = 6, units = "in")
```

**Interpretation.** Thermal sensation has weak correlation with both of the other measured indicators because it signifies actual sensation and not comfort/discomfort. On the other hand, it makes sense that TC and TA have strong negative correlation, because a high thermal comfort value indicates discomfort and thus low thermal acceptability. The strongest correlation, with a coefficient of 1, is of operative temperature with PPD and is shown in @fig-q4_addl below; this can possibly be explained by the fact that operative temperature is one of the parameters used to calculate PMV (relative humidity is also a factor in the PMV calculation and has a middling correlation value of 0.52). Another interesting correlation is of PPD and PMV: PPD is directly calculated from PMV, but the PMV-PPD model is not linear so this analysis shows a weak correlation.

```{r fig-q4_addl}
#| fig-cap: "Correlation matrix - additional"
#| fig-width: 8
#| fig-height: 4

plt_q4_strong = dat |>
   ggplot(aes(T_op, PMV)) +
      geom_point() +
      geom_smooth(method = "lm") +
      labs(x = "Operative Temperature (T_op, °C)", y = "PMV") +
      annotate("text", x = 20, y = Inf, hjust = 0, vjust = 2, label = "Correlation coefficient = 1") +
      scale_y_continuous(breaks = seq(-3, 3, 1), limits = c(-3, 3))

plt_q4_weak = dat |>
   ggplot(aes(PMV, PPD)) +
      geom_point() +
      geom_smooth() +
      labs(x = "PMV", y = "PPD (%)") +
      annotate("text", x = -3, y = Inf, hjust = 0, vjust = 2, label = "Correlation coefficient = 0.36") +
      scale_x_continuous(breaks = seq(-3, 3, 1), limits = c(-3, 3)) +
      scale_y_continuous(breaks = seq(0, 100, 10), limits = c(0, 100))

plt_q4_addl = plt_q4_strong + plt_q4_weak
ggsave("output/q4_correlation-additional.png", width = 8, height = 4, units = "in")
plt_q4_addl
```

## 5. Normality of TC and TS

```{r fig-q5}
#| fig-cap: "Normality"
#| fig-width: 8
#| fig-height: 4

q5_qqplot = function(df, var, varlabel, axlim) {
   p_val = shapiro.test(df[[var]])$p.value
   p_val = signif(p_val, digits = 2)
   
   df |>
      ggplot(aes(sample = .data[[var]])) +
         stat_qq(size = 1.5) +
         stat_qq_line(color = "gray", linetype = 2) +
         labs(title = varlabel, y = "Sample", x = "Theoretical") +
         scale_x_continuous(breaks = seq(-4, 4, 1), limits = axlim) +
         scale_y_continuous(breaks = seq(-4, 4, 1), limits = axlim) +
         annotate("text", x = -Inf, y = 4, hjust = -0.1, label = str_c("p-value: ", p_val))
}
   
plt_q5_TC = q5_qqplot(dat, "TC", "Thermal Comfort (TC)", c(1, 4))
plt_q5_TS = q5_qqplot(dat, "TS", "Thermal Sensation (TS)", c(-4, 4))
plt_q5 = plt_q5_TC + plt_q5_TS

ggsave("output/q5_normality.png", width = 8, height = 4, units = "in")
plt_q5
```

**Interpretation.** [WRITE 2–3 SENTENCES. Are they normal? Does it matter, and for what?]


## 6. Does Sex significantly influence TS?

```{r fig-q6}
#| fig-cap: "Statistical difference"
#| fig-width: 6
#| fig-height: 4

t = dat |> t_test(TS ~ Sex)

anova = dat |> anova_test(TS ~ Sex)

subtitle = str_c(
   "Two groups of TS observations: M and F\n",
   "Welch's t-test p-value: ",
   signif(t$p, 3), "\n",
   "One-way ANOVA test F-value: ",
   anova$F, 
   ", p-value: ",
   anova$p
)

plt_q6 = dat |>
   ggplot() +
      geom_boxplot(aes(Sex, TS)) +
      facet_grid(. ~ Experiment) +
      labs(y = "Thermal Sensation (TS)", subtitle = subtitle) +
      scale_y_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4))

ggsave("output/q6_differences.png", width = 6, height = 4, units = "in")
plt_q6
# Then answer this, in your text: Sex is a property of a SUBJECT, and you have
# 30 subjects but 210 rows. What is your real sample size, and what does using
# 210 do to your p-value?
# Try at least two approaches and report whether they agree.
```

**Interpretation.** [WRITE 2–3 SENTENCES]

t test & anova p value > 0.05 so difference between group means is not statistically significant
anova F value is close to 1 which means not much variation / difference between the means at all


## 7. Linear regression: TS as a function of T_op

```{r fig-q7}
#| fig-cap: "Linear regression"
#| fig-width: 6
#| fig-height: 4

model = lm(TS ~ T_op, dat)
mae = mae(dat$TS, predict(model))
rmse = rmse(dat$TS, predict(model))
caption = str_c(
   "TS = ", signif(model$coefficients["T_op"], 2),
   " × T_op - ", -signif(model$coefficients["(Intercept)"], 2),
   "\nR² = ", signif(summary(model)$r.squared, 2),
   "\nMAE = ", signif(mae, 2),
   "\nRMSE = ", signif(rmse, 2)
)
neutral_temp = -model$coefficients["(Intercept)"]/model$coefficients["T_op"]

plt_q7 = dat |>
   ggplot(aes(T_op, TS)) +
      geom_point() +
      geom_smooth(method = "lm") +
      labs(y = "Thermal Sensation (TS)", x = "Operative Temperature (T_op, °C)") +
      scale_y_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4)) +
      annotate("text", label = caption, x = 20, y = 4, hjust = 0, vjust = 1)

ggsave("output/q7_linear-regression.png", width = 6, height = 4, units = "in")
plt_q7
```

```{r fig-q7_validity}
#| fig-cap: "Linear regression diagnostics"
#| fig-width: 8
#| fig-height: 4
# The four assumptions: linearity, constant variance, normal residuals,
# independent observations. Check all four and say which ones hold.

residuals = resid(model)

# normal residuals: holds
p_val_residuals = signif(shapiro.test(residuals)$p.value, digits = 2)

plt_q7_residuals_qq = tibble(r = residuals) |>
   ggplot(aes(sample = r)) +
      stat_qq(size = 1.5) +
      stat_qq_line(color = "gray", linetype = 2) +
      labs(title = "Residuals Q-Q plot", y = "Sample", x = "Theoretical") + 
      scale_x_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4)) +
      scale_y_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4)) +
      annotate("text", x = -Inf, y = 4, hjust = -0.1, label = str_c("p-value: ", p_val_residuals))

# linearity: visual check only, not perfectly linear, close
# homoscedasticity (constant variance): holds
bp_test_p_val = signif(ncvTest(model)$p, 2)

plt_q7_residuals_variance = tibble(x = fitted(model), y = residuals) |>
   ggplot(aes(x, y)) +
      geom_point() +
      geom_smooth(method = "lm") +
      # geom_hline(yintercept = 0, color = "gray", linetype = 2) +
      labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals") +
      annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5, label = str_c("BP test p-value: ", bp_test_p_val))

# not sure how to check independence

plt_q7_check = plt_q7_residuals_qq + plt_q7_residuals_variance
ggsave("output/q7_linear-regression-diagnostics.png", width = 8, height = 4, units = "in")
plt_q7_check
```

**Interpretation.** [WRITE 2–3 SENTENCES. Report the slope with units, the R²,
and the neutral temperature — the Top at which predicted TS = 0. Compare it to
what PMV predicts.]


## 8. Your own research question

**Question:** [WRITE YOUR QUESTION]

```{r q8}
dat |>
   ggplot(aes(PMV, TS)) +
   geom_point() +
   geom_smooth(method = "lm")
```

**Interpretation.** [WRITE 2–3 SENTENCES — the take-away message you would put
in a paper.]


## Session information

```{r session}
sessionInfo()
```
