---
title: "Assignment #2 — Statistical analysis of thermal comfort research"
author: "YOUR NAME"
date: today
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
---

<!--
HOW TO USE THIS FILE
1. Click "Render" in RStudio (or run: quarto render analysis.qmd).
   If it renders before you change anything, your setup works.
2. Answer the eight questions below. Replace every [WRITE ...] prompt.
3. Every number in your text must come from a code chunk, not from
   typing it in by hand. That is the whole point of Quarto.
4. Render one more time before you submit, from a fresh R session
   (Session > Restart R). If it only renders in your current session,
   it is not reproducible.
-->

```{r setup}
library(ggplot2)
library(dplyr)
library(tidyr)

theme_set(theme_minimal(base_size = 12))

load("data/arch241a2.rda")
d <- arch241a2
```

## About this dataset — read before you start

Thirty subjects each sat through seven conditions in a climate chamber in
Guangzhou, giving 30 × 7 = 210 rows.

| Variable | Meaning | Notes |
|---|---|---|
| `Top` | Operative temperature, °C | Equals air temperature here (MRT = Ta) |
| `RH` | Relative humidity, % | Two levels: 50 and 70 |
| `Sex` | F / M | 15 subjects each |
| `Subject` | Subject ID, 1–30 | Each ID appears 7 times|
| `TS` | Thermal sensation | ASHRAE scale, −4 cold … 0 neutral … +4 hot |
| `TC` | Thermal discomfort | 0 = very comfortable … 4 = very uncomfortable |
| `TA` | Thermal acceptability | Higher = more acceptable |
| `PMV` | Predicted mean vote | Calculated from the design conditions |
| `PPD` | Predicted % dissatisfied | Calculated from PMV |
| `Experiment` | Condition A–F, plus P | P is pre-conditioning, not a test condition |

```{r orient}
str(d)
```


## 1. Statistical summary of all variables

```{r q1}
# [WRITE YOUR CODE] Summarise every variable.
# Think: does a mean make sense for a factor? For a design variable like Top?
```

**Interpretation.** [WRITE 2–3 SENTENCES. What would you put in a paper?]


## 2. Histograms of TC and TS

```{r q2}
# [WRITE YOUR CODE]
```

**Interpretation.** [WRITE 2–3 SENTENCES]


## 3. Boxplots of TS vs Sex, Subject and Experiment

```{r q3}
# [WRITE YOUR CODE] Three plots.
# The Subject one is the least obvious and the most informative — look at it carefully.
```

**Interpretation.** [WRITE 2–3 SENTENCES]


## 4. Correlation matrix

Variables: PMV, PPD, RH, TA, TC, TS, Top.

```{r q4}
# [WRITE YOUR CODE]
```

**Interpretation.** [WRITE 2–3 SENTENCES]

Two of these correlations are strange. One pair is r = 1.00, and another pair
is deterministically related yet shows a weak r. Find them, plot them, and
explain what happened.


## 5. Normality of TC and TS

```{r q5}
# [WRITE YOUR CODE] A test AND a plot. Do not rely on the p-value alone.
```

**Interpretation.** [WRITE 2–3 SENTENCES. Are they normal? Does it matter, and for what?]


## 6. Does Sex significantly influence TS?

```{r q6}
# [WRITE YOUR CODE]
# Then answer this, in your text: Sex is a property of a SUBJECT, and you have
# 30 subjects but 210 rows. What is your real sample size, and what does using
# 210 do to your p-value?
# Try at least two approaches and report whether they agree.
```

**Interpretation.** [WRITE 2–3 SENTENCES]


## 7. Linear regression: TS as a function of Top

```{r q7}
# [WRITE YOUR CODE] Fit the model, then TEST ITS VALIDITY.
# The four assumptions: linearity, constant variance, normal residuals,
# independent observations. Check all four and say which ones hold.
```

**Interpretation.** [WRITE 2–3 SENTENCES. Report the slope with units, the R²,
and the neutral temperature — the Top at which predicted TS = 0. Compare it to
what PMV predicts.]


## 8. Your own research question

**Question:** [WRITE YOUR QUESTION]

```{r q8}
# [WRITE YOUR CODE]
```

**Interpretation.** [WRITE 2–3 SENTENCES — the take-away message you would put
in a paper.]


## Session information

```{r session}
sessionInfo()
```
