---
title: "Assignment #2 — Statistical analysis of thermal comfort research"
author: "Urwa Irfan"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## Setup
```{r setup-packages}
library(tidyverse)
library(ggplot2)
library(gt)
library(ggh4x)
library(patchwork)
library(ggcorrplot)
library(rstatix)
library(Metrics)
library(car)
```

```{r setup-theme}
theme_set(theme_light(base_size = 10))
theme_update(
  strip.background = element_rect(fill = "#F5F5F5", color = "grey"),
  strip.text = element_text(color = "black"),
  strip.text.y = element_text(angle = 90)
)
theme_update(panel.grid = element_line(linetype = "dotted"))
cal_palette2 <- c("#3B7EA1", "#C4820E")
cal_palette3 = c("#3b7ea1", "#b0750c", "#003262")
cal_palette4 <- c("#003262", "#b0750c", "#3b7ea1", "#fdb515")
cal_palette5 <- c("#003262", "#b0750c", "#3b7ea1", "#fdb515", "#D9661F")
load_palette = c("#010090", "#8f0300")
```

```{r setup-data}
load("data/arch241a2.rda")
d <- arch241a2
dat = tibble(d) |>
   rename(T_op = Top) |> # for clarity
   mutate(Subject = factor(Subject, levels = as.character(sort(as.numeric(levels(Subject)))))) |> # relevel so it's sorted numerically, but kept as character for plotting
   filter(Experiment != "P") # explain why
```

## About this dataset — read before you start

Thirty subjects each sat through seven conditions in a climate chamber in
Guangzhou, giving 30 × 7 = 210 rows.

| Variable | Meaning | Notes |
|---|---|---|
| `Top` | Operative temperature, °C | Equals air temperature here (MRT = Ta) |
| `RH` | Relative humidity, % | Two levels: 50 and 70 |
| `Sex` | F / M | 15 subjects each |
| `Subject` | Subject ID, 1–30 | Each ID appears 7 times|
| `TS` | Thermal sensation | ASHRAE scale, −4 cold … 0 neutral … +4 hot |
| `TC` | Thermal discomfort | 0 = very comfortable … 4 = very uncomfortable |
| `TA` | Thermal acceptability | Higher = more acceptable |
| `PMV` | Predicted mean vote | Calculated from the design conditions |
| `PPD` | Predicted % dissatisfied | Calculated from PMV |
| `Experiment` | Condition A–F, plus P | P is pre-conditioning, not a test condition |

## 1. Statistical summary of all variables

```{r tbl-q1}
#| tbl-cap: "Statistical summary"

summarize_variable = function(df, variable) {
   df |>
      # select(.data[[variable]]) |>
      summarise(
         Minimum = min(.data[[variable]]),
         `25th Quartile` = quantile(.data[[variable]], 0.25),
         Median = median(.data[[variable]]),
         `75th Quartile` = quantile(.data[[variable]], 0.75),
         Maximum = max(.data[[variable]]),
         Mean = mean(.data[[variable]]),
         `Standard Deviation` = sd(.data[[variable]])
      ) |>
      mutate(variable = variable)
}

summary_results = tibble()
variables = names(dat)[5:9]

for (var in variables) {
   summary_results = rbind(summary_results, summarize_variable(dat, var))
}
tbl_q1 = summary_results |>
   relocate(variable) |>
   mutate(variable = case_when(
      variable == "TS" ~ "Thermal Sensation (TS)",
      variable == "TC" ~ "Thermal Comfort (TC)",
      variable == "TA" ~ "Thermal Acceptability (TA)",
      variable == "PMV" ~ "Predicted Mean Vote (PMV)",
      variable == "PPD" ~ "Percentage People Dissatisfied (PPD)",
   )) |>
   gt(rowname_col = "variable") |>
   fmt_number()

tbl_q1
gtsave(tbl_q1, "output/q1_statistical-summary.html")
# Think: does a mean make sense for a factor? For a design variable like Top?
```

**Interpretation.** [WRITE 2–3 SENTENCES. What would you put in a paper?]


## 2. Histograms of TC and TS

```{r fig-q2}
#| fig-cap: "Histograms"
#| fig-width: 6
#| fig-height: 3

plt_q2 = dat |>
   select(TC, TS) |>
   pivot_longer(cols = c(TC, TS), names_to = "variable", values_to = "value") |>
   mutate(variable = case_when(
      variable == "TS" ~ "Thermal Sensation (TS)",
      variable == "TC" ~ "Thermal Comfort (TC)")
   ) |>
   ggplot(aes(value)) +
      geom_histogram(binwidth = 1, color = "black", fill = cal_palette2[1]) +
      labs(x = "Vote", y = "Number of votes") +
      scale_y_continuous(breaks = seq(0, 120, 20)) +
      facet_wrap(~ variable, scales = "free_x") +
      facetted_pos_scales(
         x = list(
            scale_x_continuous(limits = c(1, 4), expand = expansion(mult=0.2)),
            scale_x_continuous(limits = c(-4, 4), breaks = seq(-4, 4, 1))
         )
      )

plt_q2
ggsave("output/q2_histograms.png", width = 6, height = 3, units = "in")
```

**Interpretation.** [WRITE 2–3 SENTENCES]
i have simplified to single vote count rather than continuous scale


## 3. Boxplots of TS vs Sex, Subject and Experiment

```{r fig-q3}
#| fig-cap: "Box plots"
#| fig-width: 8
#| fig-height: 6

q3_boxplot = function(df, var) {
   df |>
   ggplot() +
      geom_boxplot(aes(.data[[var]], TS)) +
      labs(y = "Thermal Sensation (TS)") +
      scale_y_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4))
}

plt_q3_sex = q3_boxplot(dat, "Sex")
plt_q3_subject = q3_boxplot(dat, "Subject")
plt_q3_experiment = q3_boxplot(dat, "Experiment")
plt_q3 = (plt_q3_sex + plt_q3_experiment) / plt_q3_subject

plt_q3
ggsave("output/q3_boxplots.png", width = 8, height = 6, units = "in")
# The Subject one is the least obvious and the most informative — look at it carefully.
```

**Interpretation.** [WRITE 2–3 SENTENCES]


## 4. Correlation matrix

Variables: PMV, PPD, RH, TA, TC, TS, Top.

```{r fig-q4}
#| fig-cap: "Correlation matrix"
#| fig-width: 6
#| fig-height: 6

plt_q4 = dat |>
   select(T_op, RH, TS, TC, TA, PMV, PPD) |>
   cor() |>
   ggcorrplot(type = "lower", outline.color = "black", lab = TRUE) +
      labs(x = "", y = "", fill = "Correlation\ncoefficient") +
      scale_fill_gradient2(low = "#4c59ea", high = "#ea544c") +
      theme_minimal() +
      theme(legend.position = "inside", legend.position.inside = c(0.2, 0.8), legend.background = element_rect(colour = "#e0e0e0"),
         axis.text.x = element_text(angle = 0))

plt_q4
ggsave("output/q4_correlation-matrix.png", width = 6, height = 6, units = "in")
```

```{r fig-q4_high-corr}
#| fig-cap: "Correlation matrix"
#| fig-width: 6
#| fig-height: 4

dat |>
   ggplot() +
      geom_point(aes(T_op, PMV))

```

PMV/PPD were directly calculated from the environmental conditions.(it just so happens that PMV is directly proportional to T_op - check?)

**Interpretation.** [WRITE 2–3 SENTENCES]

Two of these correlations are strange. One pair is r = 1.00, and another pair
is deterministically related yet shows a weak r. Find them, plot them, and
explain what happened.


## 5. Normality of TC and TS

```{r fig-q5}
#| fig-cap: "Normality"
#| fig-width: 8
#| fig-height: 4

q5_qqplot = function(df, var, varlabel, axlim) {
   p_val = shapiro.test(df[[var]])$p.value
   p_val = signif(p_val, digits = 2)
   
   df |>
      ggplot(aes(sample = .data[[var]])) +
         stat_qq(size = 1.5) +
         stat_qq_line(color = "gray", linetype = 2) +
         labs(title = varlabel, y = "Sample", x = "Theoretical") +
         scale_x_continuous(breaks = seq(-4, 4, 1), limits = axlim) +
         scale_y_continuous(breaks = seq(-4, 4, 1), limits = axlim) +
         annotate("text", x = -Inf, y = 4, hjust = -0.1, label = str_c("p-value: ", p_val))
}
   
plt_q5_TC = q5_qqplot(dat, "TC", "Thermal Comfort (TC)", c(1, 4))
plt_q5_TS = q5_qqplot(dat, "TS", "Thermal Sensation (TS)", c(-4, 4))
plt_q5 = plt_q5_TC + plt_q5_TS

ggsave("output/q5_normality.png", width = 8, height = 4, units = "in")
plt_q5
```

**Interpretation.** [WRITE 2–3 SENTENCES. Are they normal? Does it matter, and for what?]


## 6. Does Sex significantly influence TS?

```{r fig-q6}
#| fig-cap: "Statistical difference"
#| fig-width: 6
#| fig-height: 4

t = dat |> t_test(TS ~ Sex)

anova = dat |> anova_test(TS ~ Sex)

subtitle = str_c(
   "Two groups of TS observations: M and F\n",
   "Welch's t-test p-value: ",
   signif(t$p, 3), "\n",
   "One-way ANOVA test F-value: ",
   anova$F, 
   ", p-value: ",
   anova$p
)

plt_q6 = dat |>
   ggplot() +
      geom_boxplot(aes(Sex, TS)) +
      facet_grid(. ~ Experiment) +
      labs(y = "Thermal Sensation (TS)", subtitle = subtitle) +
      scale_y_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4))

ggsave("output/q6_differences.png", width = 6, height = 4, units = "in")
plt_q6
# Then answer this, in your text: Sex is a property of a SUBJECT, and you have
# 30 subjects but 210 rows. What is your real sample size, and what does using
# 210 do to your p-value?
# Try at least two approaches and report whether they agree.
```

**Interpretation.** [WRITE 2–3 SENTENCES]

t test & anova p value > 0.05 so difference between group means is not statistically significant
anova F value is close to 1 which means not much variation / difference between the means at all


## 7. Linear regression: TS as a function of T_op

```{r fig-q7}
#| fig-cap: "Linear regression"
#| fig-width: 6
#| fig-height: 4

model = lm(TS ~ T_op, dat)
mae = mae(dat$TS, predict(model))
rmse = rmse(dat$TS, predict(model))
caption = str_c(
   "TS = ", signif(model$coefficients["T_op"], 2),
   " × T_op - ", -signif(model$coefficients["(Intercept)"], 2),
   "\nR² = ", signif(summary(model)$r.squared, 2),
   "\nMAE = ", signif(mae, 2),
   "\nRMSE = ", signif(rmse, 2)
)
neutral_temp = -model$coefficients["(Intercept)"]/model$coefficients["T_op"]

plt_q7 = dat |>
   ggplot(aes(T_op, TS)) +
      geom_point() +
      geom_smooth(method = "lm", color = "black") +
      labs(y = "Thermal Sensation (TS)", x = "Operative Temperature (T_op, °C)") +
      scale_y_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4)) +
      annotate("text", label = caption, x = 20, y = 4, hjust = 0, vjust = 1)

ggsave("output/q7_linear-regression.png", width = 6, height = 4, units = "in")
plt_q7
```

```{r fig-q7_validity}
#| fig-cap: "Linear regression diagnostics"
#| fig-width: 8
#| fig-height: 4
# The four assumptions: linearity, constant variance, normal residuals,
# independent observations. Check all four and say which ones hold.

residuals = resid(model)

# normal residuals: holds
p_val_residuals = signif(shapiro.test(residuals)$p.value, digits = 2)

plt_q7_residuals_qq = tibble(r = residuals) |>
   ggplot(aes(sample = r)) +
      stat_qq(size = 1.5) +
      stat_qq_line(color = "gray", linetype = 2) +
      labs(title = "Residuals Q-Q plot", y = "Sample", x = "Theoretical") + 
      scale_x_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4)) +
      scale_y_continuous(breaks = seq(-4, 4, 1), limits = c(-4, 4)) +
      annotate("text", x = -Inf, y = 4, hjust = -0.1, label = str_c("p-value: ", p_val_residuals))

# linearity: visual check only, not perfectly linear, close
# homoscedasticity (constant variance): holds
bp_test_p_val = signif(ncvTest(model)$p, 2)

plt_q7_residuals_variance = tibble(x = fitted(model), y = residuals) |>
   ggplot(aes(x, y)) +
      geom_point() +
      geom_smooth(color = "black") +
      geom_hline(yintercept = 0, color = "gray", linetype = 2) +
      labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals") +
      annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5, label = str_c("BP test p-value: ", bp_test_p_val))

# not sure how to check independence

plt_q7_check = plt_q7_residuals_qq + plt_q7_residuals_variance
ggsave("output/q7_linear-regression-diagnostics.png", width = 8, height = 4, units = "in")
plt_q7_check
```

**Interpretation.** [WRITE 2–3 SENTENCES. Report the slope with units, the R²,
and the neutral temperature — the Top at which predicted TS = 0. Compare it to
what PMV predicts.]


## 8. Your own research question

**Question:** [WRITE YOUR QUESTION]

```{r q8}
# [WRITE YOUR CODE]
```

**Interpretation.** [WRITE 2–3 SENTENCES — the take-away message you would put
in a paper.]


## Session information

```{r session}
sessionInfo()
```
